### 2023-01-14 draft.tex
Deploying to gh-pages from @ gottacatchenall/ms_false_negatives@248cde172ef469d2562cfb1e543110f64eef9496 🚀
Deploying to gh-pages from @ gottacatchenall/ms_false_negatives@fc540949f9b43d58c34c27f827e5607a20741112 🚀
Deploying to gh-pages from @ gottacatchenall/ms_false_negatives@7aa656d41d88975532ffd237498f5e8ea2810b9f 🚀
Deploying to gh-pages from @ gottacatchenall/ms_false_negatives@9b70121eb438d5e5abf30cfe4042a4c7f958fe22 🚀

%
\href{https://orcid.org/0000-0002-6506-6487}{Michael D.\,Catchen}%
%
\,\textsuperscript{1,2}\quad %
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{3,2}\quad %
\href{https://orcid.org/0000-0002-6004-4027}{Laura\,Pollock}%
%
\,\textsuperscript{1,2}\quad %
\href{https://orcid.org/0000-0001-6075-8081}{Andrew\,Gonzalez}%
%
\,\textsuperscript{1,2}

\textsuperscript{1}\,McGill University\quad \textsuperscript{2}\,Québec
Centre for Biodiversity Sciences\quad \textsuperscript{3}\,Université de
Montréal

@@

        {\bfseries Abstract:}\,Ecosystems are composed of networks of
interacting species. These interactions allow communities of species to
persist through time through both neutral and adaptive processes. Still
a robust understanding of (and ability to predict and forecast)
interactions among species remains elusive. This knowledge-gap is
largely driven by a shortfall of data---although species occurrence data
has rapidly increased in the last decade, species interaction data has
not kept pace, largely due to the intrinsic difficulty and effort
required to sample interactions. These sampling challenges bias data and
hinder inferences about the structure and dynamics of interactions
networks. Here, we demonstrate the realized false-negative rate (the
percentage of species that actually interact but for which we do not yet
have a record) can be quite high, even in thoroughly sampled systems,
due to the intrinsic variation in abundances across species in a
community. We illustrate how a null model of occurrence detection can be
used to estimate the false-negative rate in a given dataset. One
hypothesis is that interactions between ``rare'' species are themselves
rare because these species are less likely to encounter one-another than
species of higher relative abundance. However, we demonstrate that
across several datasets of spatial or temporally replicated networks,
there are positive associations that suggest these interactions actually
exist but just are not observed. Finally, we assess how false negatives
influence various models of network prediction, and recommend directly
accounting for observation error in predictive models. We conclude by
discussing how the understanding of false-negatives can inform how we
design monitoring schemes for species interactions.\\%

@@

To simulate the process of observation, for an ecological network \(M\)
with \(S\) species, we sample abundances for each species from a
standard-log-normal distribution. For each true interaction in the
adjacency matrix \(M\) (i.e. \(M_{ij}=1\)) we estimate the probability
of observing both species \(i\) and \(j\) at a given place and time by
simulating \(n\) observations of all individuals of any a species, where
the species of the individual observed at the \(\{1,2,\dots,n\}\)-th
observation is drawn from the generated log-normal distribution of
abundances. For each pair of species \((i,j)\), if both \(i\) and \(j\)
are observed within the n-observations, the interaction is tallied as a
true positive if \(M_{ij}=1\). If only one of \(i\) or \(j\) are
observed---but not both---in these \(n\) observations, but \(M_{ij}=1\),
this is counted as a false-negative, and a true-negative otherwise. For
each pair of species \((i,j)\), if both \(i\) and \(j\) are observed
within the n-observations, the interaction is tallied as a true positive
if \(M_{ij}=1\). If only one of \(i\) or \(j\) are observed---but not
both---in these \(n\) observations, but \(M_{ij}=1\), this is counted as
a false-negative, and a true-negative otherwise (\(M_{ij} = 0\)).

@@

Here we show how to incorporate imperfect observation (both
false-negatives and false-positives) into model predictions of
interaction probability. Models for interaction prediction typically
yield a probability of interaction between each pair of species. When
these are considered with uncertainty, it is usually model-uncertainty,
e.g.~the variance in the interaction probability prediction across
several cross-validation folds, where the data is split into training
and test sets several times. Here we introduce a method for resampling
interaction probabilities (outlined in figure
fig.~\ref{fig:resampling_concept}) that simulates the observation
process with prior estimates of both false-negative and false-positive
probabilities to produce an output distribution of interaction
probabilities that accounts for observation error. We implement this in
the software package InteractionUncertaintySampler.jl.

@@

We do this by using the process for simulating a distribution of null
false-negative rates for a given dataset as described above
(fig.~\ref{fig:resampling_concept} (A)). We then consider the output
prediction from an arbitrary prediction model, which is the probability
\(p_{ij}\) that two species \(i\) and \(j\) interact. To get an estimate
of \(p_{ij}\) that accounts for observation error, we resample the
probability of each interaction \(p_{ij}\) by simulating a series of
particles. Each particle is the product of the resampling algorithm
(``Resampling'' within fig.~\ref{fig:resampling_concept} (B)), where the
value of a particle is first drawn from a Bernoulli with weight
\(p_{ij}\), and if that value is true, the value remains true with
probability \(1-p_{fn}\), and if the first draw is false, the value
remains false with probability \(1-p_{fp}\). Over many samples of
particles, the resulting frequency of `true' outcomes is a single
resample of the interaction probability \(p_{ij}^*\). Across many
resamples, this forms a distribution of probabilities which are adjusted
by the true and false negative rates.

There is also an analytic way to represent approximate this distribution
using the normal approximation to binomial. As a reminder, as the total
number of samples \(n\) from a binomial distribution with success
probability \(p\) from approaches infinity, the sum of total successes
across all samples approaches a normal distribution with mean \(np\) and
variance \(np(1-p)\). We can use this to correct the estimate \(p_{ij}\)
based on the expected false-negative-rate \(p_{fn}\) and false-positive
rate \(p_{fp}\) to obtain the limiting distribution as the number of
resamples approaches infinity for the resampled \(p_{ij}^*\) for a given
number of particles \(n_p\). We do this by first adjusting for the rates
of observation error to get the mean resampled probability,
\(\bar{p_{ij}^*}\), as

\[\bar{p_{ij}^*} = p_{ij}(1+p_{fp})(1-p_{fn})\]

@@

\[\mathcal{N}\bigg(n_p \bar{p_{ij}^*}, \sqrt{n_p \bar{p_{ij}^*} (1-\bar{p_{ij}^*})}\bigg)\]

@@

\begin{equation}\protect\hypertarget{eq:eq1}{}{\mathcal{N}\bigg(\bar{p^*_{ij}},
\sqrt{\frac{\bar{p_{ij}^*}(1-\bar{p_{ij}^*})}{n_p}} \bigg)}\label{eq:eq1}\end{equation}

We can then further truncate to remain on the interval \((0,1)\) (as the
output is a probability, although in practice often the probability mass
outside \((0,1)\) is exteremly low.

Why is this useful? Less computation, number of particles \(n_p\) as a
uncertainty width, nautral analogue to number of observations of
co-occurrence. More about this here.

As an example case study, we use a boosted-regresssion-tree to predict
interactions in a host-parasite network (Hadfield \emph{et al.} 2014)
(with features derived in the same manner as Strydom \emph{et al.}
(2021) derives features on this data) to produce a set of interaction
predictions. We then applied this method to a set of a few resampled
interaction probabilities between mammals and parasite species shown in
figure fig.~\ref{fig:resampling_concept} (C). Here we implement a simple
resampling algorithm, but this could be extended, for example by
adjusting the expected FNR for each pair of species by the relative
abundance of each species. We have implemented a general framework for
this resampling methodology in the Julia package
InteractionUncertaintySampler.jl, which enables flexible choices for
priors on false-negative and false-positive rates.

This method allows us to more effectively represent our uncertainty
about observation error in predictions of species interactions. This
also has implications for what we mean by `uncertainty' in interaction
predictions. A model's prediction can be `uncertain' in two different
ways: the model's predictions may have high variance, or the model's
predictions may be centered around a probability of interaction of
\(0.5\), where we are the most unsure about whether this interaction
exists. Improving the incorporation of different forms of uncertainty in
probabilistic interaction predictions seems a necessary next step toward
understanding what pairs of species we know the least about, in order to
prioritize sampling to provide the most new information possible.

@@

The model above doesn't consider the possibility that there are positive
or negative associations which shift the probability of species
cooccurrence together due to their interaction (Cazelles \emph{et al.}
2016). However, here we demonstrate that the probability of observing a
false negative can be higher if there is some positive association in
the occurrence of species \(A\) and \(B\). If we denote the probability
that we observe the co-occurrence of two species \(A\) and \(B\) as
\(P(AB)\) and if there is no association between the marginal
probabilities of observing \(A\) and observing \(B\), denoted \(P(A)\)
and \(P(B)\) respectively, then the probability of observing their
co-occurrence is the product of the marginal probabilities for each
species, \(P(AB) = P(A)P(B)\). In the other case where there is some
positive strength of association between observing both \(A\) and \(B\)
because this interaction is ``important'' for each species, then the
probability of observation both \(A\) and \(B\), \(P(AB)\), is greater
than \(P(A)P(B)\) as \(P(A)\) and \(P(B)\) are not independent and
instead are positively correlated, i.e.~\(P(AB)> P(A)P(B)\). In this
case, the probability of observing a single false-negative in our naive
model from fig.~\ref{fig:geometric}(A) is \(p_{fn}= 1-P(AB)\), which due
to the above inequality implies \(p_{fn}>1-P(A)P(B)\). This indicates an
increasingly greater probability of a false negative as the strength of
association gets stronger, \(P(AB) \to P(AB) \gg P(A)P(B)\). However,
this still does not consider variation in species abundance in space and
time (Poisot \emph{et al.} 2015). If positive or negative associations
between species structure variation in the distribution of \(P(AB)\)
across space/time, then the spatial/temporal biases induced by data
collection would further impact the realized false negative rate, as the
probability of false negative would not be constant for each pair of
species across sites.

>>>>

%
\href{https://orcid.org/0000-0002-6506-6487}{Michael D.\,Catchen}%
%
\,\textsuperscript{1,2}\quad %
\href{https://orcid.org/0000-0002-0735-5184}{Timothée\,Poisot}%
%
\,\textsuperscript{3,2}\quad %
\href{https://orcid.org/0000-0002-6004-4027}{Laura\,Pollock}%
%
\,\textsuperscript{1,2}\quad %
\href{https://orcid.org/0000-0001-6075-8081}{Andrew\,Gonzalez}%
%
\,\textsuperscript{1,4}

\textsuperscript{1}\,McGill University\quad \textsuperscript{2}\,Québec
Centre for Biodiversity Sciences\quad \textsuperscript{3}\,Université de
Montréal\quad \textsuperscript{4}\,Québec Centre for Biodiversity
Science

@@

        {\bfseries Abstract:}\,Ecosystems are composed of networks of
interacting species. These interactions allow communities of species to
persist through time through both neutral and adaptive processes. Still
a robust understanding of (and ability to predict and forecast)
interactions among species remains elusive. This knowledge-gap is
largely driven by a shortfall of data---although species occurrence data
has rapidly increased in the last decade, species interaction data has
not kept pace, largely due to the intrinsic difficulty and effort
required to sample interactions. These sampling challenges bias data and
hinder inferences about the structure and dynamics of interactions
networks. Here, we demonstrate the realized false-negative rate (the
percentage of species that actually interact but for which we do not yet
have a record) can be quite high, even in thoroughly sampled systems,
due to the intrinsic variation in abundances across species in a
community. We illustrate how a null model of occurrence detection can be
used to estimate the false-negative rate in a given dataset. We also
show how to directly incorporate uncertainty due to observation error
into model-based predictions of interaction probabilities between
species. One hypothesis is that interactions between ``rare'' species
are themselves rare because these species are less likely to encounter
one-another than species of higher relative abundance. However, we
demonstrate that across several datasets of spatial or temporally
replicated networks, there are positive associations that suggest these
interactions actually exist but just are not observed. Finally, we
assess how false negatives influence various models of network
prediction, and recommend directly accounting for observation error in
predictive models. We conclude by discussing how the understanding of
false-negatives can inform how we design monitoring schemes for species
interactions.\\%

@@

To simulate the process of observation, for an ecological network \(M\)
with \(S\) species, we sample abundances for each species from a
standard-log-normal distribution. For each true interaction in the
adjacency matrix \(M\) (i.e. \(M_{ij}=1\)) we estimate the probability
of observing both species \(i\) and \(j\) at a given place and time by
simulating \(n\) observations of all individuals of any a species, where
the species of the individual observed at the \(\{1,2,\dots,n\}\)-th
observation is drawn from the generated log-normal distribution of
abundances. For each pair of species \((i,j)\), if both \(i\) and \(j\)
are observed within the n-observations, the interaction is tallied as a
true positive if \(M_{ij}=1\). If only one of \(i\) or \(j\) are
observed---but not both---in these \(n\) observations, but \(M_{ij}=1\),
this is counted as a false-negative, and a true-negative otherwise. For
each pair of species \((i,j)\), if both \(i\) and \(j\) are observed
within the n-observations, the interaction is tallied as a true positive
if \(M_{ij}=1\). If only one of \(i\) or \(j\) are observed---but not
both---in these \(n\) observations, but \(M_{ij}=1\), this is counted as
a false-negative, and a true-negative otherwise (\(M_{ij} = 0\)). This
process is illustrated conceptually in
fig.~\ref{fig:resampling_concept}(A).

@@

Here we show how to incorporate uncertainty into model predictions of
interaction probability to account for imperfect observation (both
false-negatives and false-positives). Models for interaction prediction
typically yield a probability of interaction between each pair of
species, \(p_{ij}\). When these are considered with uncertainty, it is
usually model-uncertainty, e.g.~the variance in the interaction
probability prediction across several cross-validation folds, where the
data is split into training and test sets several times. The method we
introduce adjusts the value of a model's predictions to produce a
distribution of interaction probabilities, which are adjusted by a given
false-negative-rate \(p_{fn}\) and false-positive-rate \(p_{fp}\)
(outlined in figure fig.~\ref{fig:resampling_concept}). We describe
first how to sample from this distribution of adjusted interaction
probabilities via simulation, and show that this distribution can be
well-approximated analytically.

@@

We then consider the output prediction from an arbitrary prediction
model, which is the probability \(p_{ij}\) that two species \(i\) and
\(j\) interact. To get an estimate of \(p_{ij}\) that accounts for
observation error, we resample the probability of each interaction
\(p_{ij}\) by simulating a set of several `particles,' where each
particle is a realization of an interaction occurring (either true or
false with probabilities \(p_{ij}\) and \(1-p_{ij}\) respectively) and
then being correctly observed with probabilities given by \(p_{fp}\) and
\(p_{fn}\) to yield a single boolean outcome for each particle.
(``Resampling'' within fig.~\ref{fig:resampling_concept} (B)). Over many
samples of particles, the resulting frequency of `true' outcomes is a
single resample of the interaction probability \(p_{ij}^*\). Across
several samples each of several particles, this forms a distribution of
probabilities which are adjusted by the true and false negative rates.

There is also an analytic way to approximate this distribution using the
normal approximation to binomial. As a reminder, as the total number of
samples \(n\) from a binomial distribution with success probability
\(p\) from approaches infinity, the sum of total successes across all
samples approaches a normal distribution with mean \(np\) and variance
\(np(1-p)\). We can use this to correct the estimate \(p_{ij}\) based on
the expected false-negative-rate \(p_{fn}\) and false-positive rate
\(p_{fp}\) to obtain the limiting distribution as the number of
resamples approaches infinity for the resampled \(p_{ij}^*\) for a given
number of particles \(n_p\). We do this by first adjusting for the rates
of observation error to get the mean resampled probability,
\(\mathbb{E}[{p_{ij}^*}]\), as

\[
\mathbb{E}[{p_{ij}^*}] = p_{ij}(1-p_{fp})+ (1-p_{ij})p_{fn}
\]

@@

\[
\sum_{i=1}^{n_p} p_{ij}^* \sim \mathcal{N}\bigg(n_p \cdot \mathbb{E}[p_{ij}^*], \sqrt{n_p\mathbb{E}[p_{ij}^*] (1- \mathbb{E}[p_{ij}^*])}\bigg)
\]

@@

\begin{equation}\protect\hypertarget{eq:eq1}{}{p_{ij}^* \sim \mathcal{N}\bigg( \mathbb{E}[p_{ij}^*] , \sqrt{\frac{\mathbb{E}[p_{ij}^*]
(1-\mathbb{E}[p_{ij}^*] )}{n_p}} \bigg)}\label{eq:eq1}\end{equation}

We can then further truncate to remain on the interval \((0,1)\) (as the
output is a probability, although in practice often the probability mass
outside \((0,1)\) is exteremly low. As an example case study, we use a
boosted-regression-tree to predict interactions in a host-parasite
network (Hadfield \emph{et al.} 2014) (with features derived in the same
manner as Strydom \emph{et al.} (2021) derives features on this data) to
produce a set of interaction predictions. We then applied this method to
a set of a few resampled interaction probabilities between mammals and
parasite species shown in figure fig.~\ref{fig:resampling_concept}(C).

Why is this useful? For one, this analytic method avoids the extra
computation required by simulating samples from this distribution
directly. Further, it enables continuous exanination of the number of
particles \(n_p\) as a uncertainty width. The natural analogue for the
number of particles sampled is the number of observations of
co-occurrence for a given pair of species---the fewer the particles, the
higher the variance of the resulting approximation (see supplemental
figure 1 for an example). Equation (\textbf{ref:eq:eq1?}) is undefined
for 0 particles (i.e.~0 observations co-occurrence), although as \(n_p\)
approaches \(0\) the approxated normal (once truncated) approaches a
uniform distribution on the interval \((0,1)\), the maximum entropy
distribution where we have no information about the possibility of an
interaction.

This also has implications for what we mean by `uncertainty' in
interaction predictions. A model's prediction can be `uncertain' in two
different ways: (1) the model's predictions may have high variance, or
(2) the model's predictions may be centered around a probability of
interaction of \(0.5\), where we are the most unsure about whether this
interaction exists . For non-zero \(p_{fn}\) and \(p_{fp}\), the above
model always moves the mean interaction probability toward 0.5
(supplemental figure 2), siding toward higher uncertainty in the latter
(higher entropy) sense. Improving the incorporation of different forms
of uncertainty in probabilistic interaction predictions seems a
necessary next step toward understanding what pairs of species we know
the least about, in order to prioritize sampling to provide the most new
information possible.

@@

The model above doesn't consider the possibility that there are positive
or negative associations which shift the probability of species
cooccurrence together due to their interaction (Cazelles \emph{et al.}
2016). However, here we demonstrate that the probability of observing a
false-negative can be higher if there is some positive association in
the occurrence of species \(A\) and \(B\). If we denote the probability
that we observe the co-occurrence of two species \(A\) and \(B\) as
\(P(AB)\) and if there is no association between the marginal
probabilities of observing \(A\) and observing \(B\), denoted \(P(A)\)
and \(P(B)\) respectively, then the probability of observing their
co-occurrence is the product of the marginal probabilities for each
species, \(P(AB) = P(A)P(B)\). In the other case where there is some
positive strength of association between observing both \(A\) and \(B\)
because this interaction is ``important'' for each species, then the
probability of observation both \(A\) and \(B\), \(P(AB)\), is greater
than \(P(A)P(B)\) as \(P(A)\) and \(P(B)\) are not independent and
instead are positively correlated, i.e.~\(P(AB)> P(A)P(B)\). In this
case, the probability of observing a single false-negative in our naive
model from fig.~\ref{fig:geometric}(A) is \(p_{fn}= 1-P(AB)\), which due
to the above inequality implies \(p_{fn}>1-P(A)P(B)\). This indicates an
increasingly greater probability of a false negative as the strength of
association gets stronger, \(P(AB) \to P(AB) \gg P(A)P(B)\). However,
this still does not consider variation in species abundance in space and
time (Poisot \emph{et al.} 2015). If positive or negative associations
between species structure variation in the distribution of \(P(AB)\)
across space/time, then the spatial/temporal biases induced by data
collection would further impact the realized false negative rate, as the
probability of false negative would not be constant for each pair of
species across sites.
